Chasing \rainbows{} with the Habitable Worlds Observatory

### 2023-02-24 src/tex/ms.tex
[showyourwork] overleaf sync

% Chasing rainbows with the Habitable Worlds Observatory
\title{Chasing \rainbows \hspace{1pt} with the Habitable Worlds Observatory}
%Scattering phase angle probed by directly imaged planets, or
%Don't ask us about contrast

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{1,}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{2,9}$,
    Sarah L. Casewell$^{3}$,
    Nicolas B. Cowan$^{4}$,
    David Doelman$^{5}$,\newauthor
    Timothy D. Gebhard$^{6}$,
    Matthew Kenworthy$^{5}$,
    Johan Mazoyer$^{7}$,
    Maxwell A. Millar-Blanchaer$^{8}$
    \newauthor [remaining workshop participants in alphabetical order] \newauthor 
    \\
    $^{1}$ Department of Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH, UK\\
    $^{2}$ Department of Earth and Planetary Sciences, University of California, Riverside, CA 92521, USA \\
    $^{3}$ Centre for Exoplanet Research, School of Physics and Astronomy, University of Leicester, University Road, Leicester, LE1 7RH, UK\\
    $^{4}$ Department of Earth \& Planetary Sciences and Department of Physics, McGill University, 3600 rue University, Montréal, QC, H3A 2T8, Canada\\
    $^{5}$ Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\\
    $^{6}$ Max Planck Institute for Intelligent Systems, Max Planck Ring 4, 72076 Tübingen, Germany \& ETH Zurich, Switzerland \\
    $^{7}$ LESIA, Observatoire de Paris, Université PSL, CNRS, Sorbonne Université, Université de Paris, F-92195 Meudon, France \\
    $^{8}$ Department of Physics, University of California, Santa Barbara, CA 93106, USA 
    $^{9}$ NASA Nexus for Exoplanet System Science, Virtual Planetary Laboratory Team, Seattle, WA 98195, USA
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory, a coronagraphic mission which would detect rocky planets in the habitable zone, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability. Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. The glint signature at scattering phases of $\sim$50--70$^\circ$ would be accessible to the Habitable Worlds Observatory in XX systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in YY systems. The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with IWA.

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability for about 25 habitable zone planets. This assertion requires an instrument with a coronograph capable of high contrast imaging in the optical-near infrared. Following this release, NASA recently announced the start of the development of the Habitable Worlds Observatory (HWO). The precursor technology recommended by the Decadal Survey also lists "direct imaging to probe polarized ocean glint on terrestrial planets" as one of the priority capabilities (\citet[Box E.1 in][]{decadal}) for ground and space-based observatories.
The performance and precise characteristics of this telescope (on-axis or off-axis, exact diameter, type of segmentation, type of coronagraph) are still to be determined. The development will be heavily influenced by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.  The Inner Working angle (\IWA) of the resulting telescope will be dependent on these choices, which in turn will have a significant impact on the expected exo-Earth yield \cite{Stark2019_exoplanetyield}.

Extreme Coronagraph for Living Planetary Systems (ECLIPS) coronograph for LUVOIR was a proposed \SIrange{200}{2000}{\nano\meter} instrument for exoplanet characterisation, with a similar coronagraph operating between \num{450} and \SI{1800}{\nano\meter} planned for HabEx. 
POLLUX, a proposed UV instrument for the LUVOIR study could also allow the detection of polarised light from hot Jupiters at \SI{300}{\nano\meter} \citep{Bouret2018_pollux}.  However, the challenges associated with the design of a coronagraph in UV are enormous, and it is unlikely that the HWO will include a UV high-contrast instrument sensitive to Earth-like planets.

@@

 Phase curves of planets are more complex than simple Lambertian curves. 
 The phase curve of an edge on system in unpolarized light is dominated by the reflectance and backscatter of the surface, clouds, and atmosphere. Though this behaviour is not Lambertian, the planet will, in general, get brighter at fuller phases (small scattering angles) with a peak at full phase from cloud backscatter \citet[see, for example,][]{kopparla2018}. In cases where surface liquid (sea) is present, the specular reflection can create a notable addition to the phase curve at broad scattering angles near crescent phase \citep{Robinson_2010}. Exoplanet phase curves can also contain a rainbow signature at gibbous phases (moderately narrow scattering angles), though this is a small feature in comparison to the reflectance.

@@

\begin{figure*}
    \centering
    \includegraphics{figures/scatterplot.pdf}
    \caption{
        Scatter plot at 3 $\lambda/D$ for the target sample showing stellar effective temperature and stellar distance. 
        The size of the points represents the angular separation of the star and planet in milliarcseconds as presented in the target list. The colour of the points shows the atmospheric phenomenon that can be detected with darker colors including all lighter (yellow) color phenomenon. Thus dark blue points are systems which have the most key features, as systems in which the angles required to see the rainbow are probed will also have the angles required to see the Rayleigh scattering probed.
    }
    \label{fig:scatterplot}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/madness.pdf}  
    \caption{
        A example of the eccentric orbits generated for the stellar sample. The orbits are scaled by the Earth-equivalent flux distance. The shaded region indicates \IWA's of 1, 2, 3 and 4~$\lambda / D$ corresponding to 21, 42, 63 and 84 mas. The \IWA can significantly affect the range of scattering phases observable with each orbit.
    }
    \label{fig:ball-o-yarn}
\end{figure*}

>>>>

% Chasing rainbows with the Habitable Worlds Observatory
\title{Chasing \rainbows{} with the Habitable Worlds Observatory}
%Scattering phase angle probed by directly imaged planets, or
%Don't ask us about contrast

% Authors are in alphabetical order:
\author[Sophia R. Vaughan et al.]{%
    Sophia R. Vaughan$^{1,}$\thanks{Correspondence:  \url{sophia.vaughan@physics.ox.ac.uk}}
    Kimberly Bott$^{2,9}$,
    Sarah L. Casewell$^{3}$,
    Nicolas B. Cowan$^{4}$,
    David Doelman$^{5}$,\newauthor
    Timothy D. Gebhard$^{6}$,
    Matthew Kenworthy$^{5}$,
    Johan Mazoyer$^{7}$,
    Maxwell A. Millar-Blanchaer$^{8}$
    \newauthor [remaining workshop participants in alphabetical order] \newauthor 
    \\
    $^{1}$ Department of Astrophysics, University of Oxford, Denys Wilkinson Building, Keble Road, Oxford OX1 3RH, UK\\
    $^{2}$ Department of Earth and Planetary Sciences, University of California, Riverside, CA 92521, USA \\
    $^{3}$ Centre for Exoplanet Research, School of Physics and Astronomy, University of Leicester, University Road, Leicester, LE1 7RH, UK\\
    $^{4}$ Department of Earth \& Planetary Sciences and Department of Physics, McGill University, 3600 rue University, Montréal, QC, H3A 2T8, Canada\\
    $^{5}$ Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\\
    $^{6}$ Max Planck Institute for Intelligent Systems, Max Planck Ring 4, 72076 Tübingen, Germany \& ETH Zurich, Switzerland \\
    $^{7}$ LESIA, Observatoire de Paris, Université PSL, CNRS, Sorbonne Université, Université de Paris, F-92195 Meudon, France \\
    $^{8}$ Department of Physics, University of California, Santa Barbara, CA 93106, USA \\
    $^{9}$ NASA Nexus for Exoplanet System Science, Virtual Planetary Laboratory Team, Seattle, WA 98195, USA
}
\date{Draft version: \today}

@@

\begin{abstract}
NASA recently announced the Habitable Worlds Observatory, a coronagraphic mission which would detect rocky planets in the habitable zone, establish their habitability, and search them for biosignatures. 
%Numerous instrumental and observational challenges have still to be overcome to design a coronagraphic mission with that goal. After a research phase aiming at the detection of a few tens of exo-Earth in the habitable zone, the Habitable Worlds Observatory will use its spectroscopic and polarimetric capabilities to probe these planets' atmosphere and surface properties. 
Surface liquid water is central to the definition of planetary habitability.
%
Photometric and polarimetric phase variations can show the presence of specularly reflecting oceans on exoplanets, in addition to being sensitive to rainbows and other cloud scattering effects. Direct imaging missions are optimised to detect planets near quadrature, which may obscure some of the phases at which these features are strongest. The range of scattering phases accessible for an exoplanet can be limited by its orbital inclination or the coronagraph's inner working angle. Planets that orbit close to edge-on have accessible phase angles limited by obscuration by the coronagraph. We use the list of target stars for the Habitable Worlds Observatory to estimate the number of exo-Earths that could be searched for non-Lambertian scattering phenomena. We find that exo-Earths in systems listed in this Habitable Worlds Observatory catalog will have a detectable Rayleigh scattering peak. The glint signature at scattering phases of $\sim$50--70$^\circ$ would be accessible to the Habitable Worlds Observatory in XX systems, and the rainbow of water clouds at phases of 140--160$^\circ$ would be accessible in YY systems. The number of systems in which an exo-Earth could be searched for these scattering phenomena increases with orbital eccentricity and scales approximately inversely with IWA.

@@

The National Academy of Sciences Astronomy \& Astrophysics 2020 Decadal Survey \citep{decadal} recommended the first in a new \enquote{Great Observatories} program a telescope with the capability to detect signatures of habitability for about 25 habitable zone planets.
%
This assertion requires an instrument with a coronograph capable of high contrast imaging in the optical-near infrared.
%
Following this release, NASA recently announced the start of the development of the Habitable Worlds Observatory (HWO).
%
The precursor technology recommended by the Decadal Survey also lists ``direct imaging to probe polarized ocean glint on terrestrial planets'' as one of the priority capabilities \citep[Box E.1 in][]{decadal} for ground and space-based observatories.
%
The performance and precise characteristics of this telescope (on-axis or off-axis, exact diameter, type of segmentation, type of coronagraph) are still to be determined.
%
The development will be heavily influenced by the LUVOIR \citep{LUVOIR2019} and HabEx \citep{HabEx_2020} preparatory studies.
%
The Inner Working angle (\IWA) of the resulting telescope will be dependent on these choices, which in turn will have a significant impact on the expected exo-Earth yield \citep{Stark2019_exoplanetyield}.

Extreme Coronagraph for Living Planetary Systems (ECLIPS) coronograph for LUVOIR was a proposed \SIrange{200}{2000}{\nano\meter} instrument for exoplanet characterisation, with a similar coronagraph operating between \num{450} and \SI{1800}{\nano\meter} planned for HabEx.
%
POLLUX, a proposed UV instrument for the LUVOIR study could also allow the detection of polarised light from hot Jupiters at \SI{300}{\nano\meter} \citep{Bouret2018_pollux}.
%
However, the challenges associated with the design of a coronagraph in UV are enormous, and it is unlikely that the HWO will include a UV high-contrast instrument sensitive to Earth-like planets.

@@

 Phase curves of planets are more complex than simple Lambertian curves.
 %
 The phase curve of an edge on system in unpolarized light is dominated by the reflectance and backscatter of the surface, clouds, and atmosphere.
 %
 Though this behaviour is not Lambertian, the planet will, in general, get brighter at fuller phases (small scattering angles) with a peak at full phase from cloud backscatter \citep[see, for example,][]{kopparla2018}.
 %
 In cases where surface liquid (sea) is present, the specular reflection can create a notable addition to the phase curve at broad scattering angles near crescent phase \citep{Robinson_2010}.
 %
 Exoplanet phase curves can also contain a rainbow signature at gibbous phases (moderately narrow scattering angles), though this is a small feature in comparison to the reflectance.

@@

\begin{figure*}
    \centering
    \includegraphics{figures/scatterplot.pdf}
    \caption{
        Scatter plot at 3 $\lambda/D$ for the target sample showing stellar effective temperature and stellar distance. 
        The size of the points represents the angular separation of the star and planet in milliarcseconds as presented in the target list. The colour of the points shows the atmospheric phenomenon that can be detected with darker colors including all lighter (yellow) color phenomenon. Thus dark blue points are systems which have the most key features, as systems in which the angles required to see the rainbow are probed will also have the angles required to see the Rayleigh scattering probed.
    }
    \label{fig:scatterplot}
    \script{create-scatterplot.py}
\end{figure*}

@@

\begin{figure*}
    \centering
    \includegraphics[width=\textwidth]{figures/ball_of_yarn.pdf}  
    \caption{
        A example of the eccentric orbits generated for the stellar sample. The orbits are scaled by the Earth-equivalent flux distance. The shaded region indicates \IWA's of 1, 2, 3 and 4~$\lambda / D$ corresponding to 21, 42, 63 and 84 mas. The \IWA can significantly affect the range of scattering phases observable with each orbit.
    }
    \label{fig:ball-o-yarn}
    \script{make_yarn_plot.py}
\end{figure*}
